---
title: "DESeq2 Analysis: CN Exponential vs Stationary"
author: "Jan Dolinsek & https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html & AI"
date: "08.07.2026"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: default
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Installation and Setup

## Install BiocManager

```{r install-biocmanager, eval=FALSE}
install.packages("BiocManager")
```

## Install Necessary Packages

```{r install-packages, eval=FALSE}
BiocManager::install("rtracklayer")
BiocManager::install("DESeq2")
BiocManager::install("apeglm")
BiocManager::install("ashr")
BiocManager::install("vsn")
BiocManager::install("hexbin")
BiocManager::install("IHW")
BiocManager::install("clusterProfiler")
```

## Load Libraries

```{r load-libraries}
library("DESeq2")
library("rtracklayer")
library("openxlsx")
library("pheatmap")
library("ggplot2")
library("apeglm")
library("ashr")
library("vsn")
library("hexbin")
library("RColorBrewer")
library("IHW")
```

# Data Input and Preparation

## Set Working Directory and Read Count Table
## Data Quality Checks

```{r read-count-table}
pc_folders <- c("C:/", "C:/Users/dolinsekj/")
pc_folder <- 2

setwd(paste0(pc_folders[pc_folder], "Dropbox/FHWN/M.BDSC.B.23.AA Masterarbeit/MTX/"))

samples <- c("1_SIE/featurecounts/counts_GCA_976997615.tsv",
             "1_SIE/featurecounts/counts_GCA_977007255.tsv",
             "1_SIE/featurecounts/counts_GCA_977008005.tsv",
             "1_SIE/featurecounts/counts_GCA_977008225.tsv",
             "9_SIM/featurecounts/counts_GCA_976997325.tsv",
             "9_SIM/featurecounts/counts_GCA_976997375.tsv",
             "9_SIM/featurecounts/counts_bin.3.tsv",
             "9_SIM/featurecounts/counts_bin.5.tsv",
             "9_SIM/featurecounts/counts_bin.6.tsv",
             "9_SIM/featurecounts/counts_bin.7.tsv",
             "4_STO/featurecounts/counts_GCA_976996985.tsv",
             "4_STO/featurecounts/counts_GCA_977004765.tsv",
             "4_STO/featurecounts/counts_GCA_977004785.tsv",
             "4_STO/featurecounts/counts_GCA_977004805.tsv"
             )
sample <- 9

enrichment <- strsplit(samples[sample], "/")[[1]][1]
genome <- strsplit(samples[sample], "/")[[1]][3]

dir.create(paste0(pc_folders[pc_folder],
                        "Dropbox/FHWN/M.BDSC.B.23.AA Masterarbeit/MTX/",
                        enrichment,
                        '/featurecounts/individual/',
                        genome),
                        recursive = TRUE, showWarnings = FALSE)


count_in <- paste0(pc_folders[pc_folder], "Dropbox/FHWN/M.BDSC.B.23.AA Masterarbeit/MTX/", samples[sample])
counts_raw <- read.table(count_in, header = TRUE, row.names = "Geneid")

metadata_in <- paste0(pc_folders[pc_folder],
                        "Dropbox/FHWN/M.BDSC.B.23.AA Masterarbeit/MTX/",
                        enrichment,
                        "/featurecounts/design_table.tsv")

metadata <- read.table(metadata_in, header = TRUE, sep = "\t", row.names = "samples")

rownames(metadata) <- sub("fb", "", rownames(metadata))

# Check that metadata and count table align correctly
all(rownames(metadata) %in% colnames(counts_raw)) # must be TRUE
all(rownames(metadata) == colnames(counts_raw))   # must be TRUE
```

# Create DESeqDataSet
# Prefiltering
# Initial DESeq Analysis
# Log fold change shrinkage for visualization and ranking
# Independent Hypothesis Weighting

```{r dds}
# create DESeqDataSet
dds <- DESeqDataSetFromMatrix(countData = counts_raw,
                              colData = metadata,
                              design = ~ condition + group)
dds

# do the prefiltering
smallestGroupSize <- 3
keep <- rowSums(counts(dds) >= 5) >= smallestGroupSize
dds <- dds[keep, ]

nrow(dds)

# run the deseq2
dds$condition <- factor(dds$condition, levels = c("NH4","CN"))
dds <- DESeq(dds)

as.data.frame(colData(dds))

levels(dds$condition)
levels(dds$group)

res <- results(dds, contrast = c("condition", "CN", "NH4"))
levels(dds$condition)
resultsNames(dds)

# Log fold change shrinkage for visualization and ranking
resLFC_apeglm <- lfcShrink(dds, coef="condition_CN_vs_NH4", type="apeglm")
resLFC_apeglm <- lfcShrink(dds, coef="condition_CN_vs_NH4", type="apeglm")
resLFC_ashr <- lfcShrink(dds, coef="condition_CN_vs_NH4", type="ashr")
resLFC_normal <- lfcShrink(dds, coef="condition_CN_vs_NH4", type="normal")

# # Independent hypothesis weighting
# resIHW <- results(dds, filterFun=ihw)
# summary(resIHW)
# sum(resIHW$padj < 0.05, na.rm=TRUE)
# metadata(resIHW)$ihwResult

# Export Results


# dir.create(paste0(pc_folders[pc_folder],
#                         "Dropbox/FHWN/M.BDSC.B.23.AA Masterarbeit/MTX/",
#                         enrichment,
#                         '/featurecounts/individual/',
#                         genome),
#                         recursive = TRUE, showWarnings = FALSE)

write.csv(as.data.frame(resLFC_apeglm), 
          file = paste0(pc_folders[pc_folder],
                        "Dropbox/FHWN/M.BDSC.B.23.AA Masterarbeit/MTX/",
                        enrichment,
                        '/featurecounts/individual/',
                        genome,
                        "/dds_all_CN_vs_NH4_resLFC_apeglm.csv"))


# MA Plots
plotMA(res, ylim = c(-4, 9))
plotMA(resLFC_apeglm, ylim = c(-4, 9))

# Dispersion Plot
plotDispEsts(dds)

# Count Plots
plotCounts(dds, gene = which.min(res$padj), intgroup = "condition")




vsd <- varianceStabilizingTransformation(dds, blind = FALSE)
# vsd <- vst(dds, blind=FALSE) # use this to see what happens
rld <- rlog(dds, blind=FALSE)
head(assay(vsd), 3)

ntd <- normTransform(dds)
meanSdPlot(assay(dds))
meanSdPlot(assay(ntd))
meanSdPlot(assay(vsd))
meanSdPlot(assay(rld))



# Heatmap of the count matrix ---------------------------------------------

select <- order(rowMeans(counts(dds,normalized=TRUE)),
                decreasing=TRUE)[1:20]
df <- as.data.frame(colData(dds)[,c("condition","group")])
pheatmap(assay(ntd)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df)

pheatmap(assay(vsd)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df)

pheatmap(assay(rld)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df)



# Heatmap of the sample-to-sample distances -------------------------------

sampleDists <- dist(t(assay(vsd)))

sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$group, vsd$condition, sep="-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)


# Principal component plot of the samples ---------------------------------


plotPCA(vsd, intgroup=c("condition", "group"))


pcaData <- plotPCA(vsd, intgroup=c("condition", "group"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=condition, shape=group)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()


```


## Subset Data for "CN" condition
# Perform DESeq Analysis on Subset
# Log Fold Change Shrinkage with apeglm
# Independent Hypothesis Weighting
# Export Results
# MA Plots
# Dispersion Plot
# Count Plots

```{r subset-cn}
# Subset Data for CN Condition
dds_CN_adapt_vs_exp <- dds[, dds$condition == "CN"]

dds_CN_adapt_vs_exp$group <- droplevels(dds_CN_adapt_vs_exp$group)
dds_CN_adapt_vs_exp$condition <- droplevels(dds_CN_adapt_vs_exp$condition)

dds_CN_adapt_vs_exp$group <- relevel(dds_CN_adapt_vs_exp$group, ref = "exponential")

design(dds_CN_adapt_vs_exp) <- ~ group

# Perform DESeq Analysis on Subset
dds_CN_adapt_vs_exp <- DESeq(dds_CN_adapt_vs_exp)
res_CN_adapt_vs_exp <- results(dds_CN_adapt_vs_exp, contrast = c("group", "adaptation", "exponential"))
resultsNames(dds_CN_adapt_vs_exp)

# Log Fold Change Shrinkage with apeglm
resLFC_apeglm_CN_adapt_vs_exp <- lfcShrink(dds_CN_adapt_vs_exp, 
                                           coef = "group_adaptation_vs_exponential", 
                                           type = "apeglm")

# # Independent Hypothesis Weighting
# resIHW <- results(dds_CN_adapt_vs_exp, filterFun = ihw)
# summary(resIHW)
# sum(resIHW$padj < 0.05, na.rm = TRUE)
# metadata(resIHW)$ihwResult



write.csv(as.data.frame(resLFC_apeglm_CN_adapt_vs_exp), 
          file = paste0(pc_folders[pc_folder],
                        "Dropbox/FHWN/M.BDSC.B.23.AA Masterarbeit/MTX/",
                        enrichment,
                        '/featurecounts/individual/',
                        genome,
                        "/dds_CN_adapt_vs_exp_resLFC_apeglm.csv"))


# MA Plots
plotMA(res_CN_adapt_vs_exp, ylim = c(-4, 9))
plotMA(resLFC_apeglm_CN_adapt_vs_exp, ylim = c(-4, 9))

# Dispersion Plot
plotDispEsts(dds_CN_adapt_vs_exp)

# Count Plots
plotCounts(dds_CN_adapt_vs_exp, gene = which.min(res$padj), intgroup = "group")
```

## Subset Data for "CN" condition
# Perform DESeq Analysis on Subset
# Log Fold Change Shrinkage with apeglm
# Independent Hypothesis Weighting
# Export Results
# MA Plots
# Dispersion Plot
# Count Plots

```{r subset-cn}
# Subset Data for CN Condition
dds_CN_exp_vs_stat <- dds[, dds$condition == "CN"]

dds_CN_exp_vs_stat$group <- droplevels(dds_CN_exp_vs_stat$group)
dds_CN_exp_vs_stat$condition <- droplevels(dds_CN_exp_vs_stat$condition)

dds_CN_exp_vs_stat$group <- relevel(dds_CN_exp_vs_stat$group, ref = "stationary")

design(dds_CN_exp_vs_stat) <- ~ group

# Perform DESeq Analysis on Subset
dds_CN_exp_vs_stat <- DESeq(dds_CN_exp_vs_stat)
res_CN_exp_vs_stat <- results(dds_CN_exp_vs_stat, contrast = c("group", "exponential", "stationary"))
resultsNames(dds_CN_exp_vs_stat)

# Log Fold Change Shrinkage with apeglm
resLFC_apeglm_CN_exp_vs_stat <- lfcShrink(dds_CN_exp_vs_stat, 
                                           coef = "group_exponential_vs_stationary", 
                                           type = "apeglm")

# # Independent Hypothesis Weighting
# resIHW <- results(dds_CN_exp_vs_stat, filterFun = ihw)
# summary(resIHW)
# sum(resIHW$padj < 0.05, na.rm = TRUE)
# metadata(resIHW)$ihwResult


# dir.create(paste0(pc_folders[pc_folder],
#                         "Dropbox/FHWN/M.BDSC.B.23.AA Masterarbeit/MTX/",
#                         enrichment,
#                         '/featurecounts/individual/',
#                         genome),
#                         recursive = TRUE, showWarnings = FALSE)

write.csv(as.data.frame(resLFC_apeglm_CN_exp_vs_stat), 
          file = paste0(pc_folders[pc_folder],
                        "Dropbox/FHWN/M.BDSC.B.23.AA Masterarbeit/MTX/",
                        enrichment,
                        '/featurecounts/individual/',
                        genome,
                        "/dds_CN_exp_vs_stat_resLFC_apeglm.csv"))


# MA Plots
plotMA(res_CN_exp_vs_stat, ylim = c(-4, 9))
plotMA(resLFC_apeglm_CN_exp_vs_stat, ylim = c(-4, 9))

# Dispersion Plot
plotDispEsts(dds_CN_exp_vs_stat)

# Count Plots
plotCounts(dds_CN_exp_vs_stat, gene = which.min(res$padj), intgroup = "group")
```

## Subset Data for "CN" condition
# Perform DESeq Analysis on Subset
# Log Fold Change Shrinkage with apeglm
# Independent Hypothesis Weighting
# Export Results
# MA Plots
# Dispersion Plot
# Count Plots

```{r subset-cn}
# Subset Data for CN Condition
dds_CN_adapt_vs_stat <- dds[, dds$condition == "CN"]

dds_CN_adapt_vs_stat$group <- droplevels(dds_CN_adapt_vs_stat$group)
dds_CN_adapt_vs_stat$condition <- droplevels(dds_CN_adapt_vs_stat$condition)

dds_CN_adapt_vs_stat$group <- relevel(dds_CN_adapt_vs_stat$group, ref = "stationary")

design(dds_CN_adapt_vs_stat) <- ~ group

# Perform DESeq Analysis on Subset
dds_CN_adapt_vs_stat <- DESeq(dds_CN_adapt_vs_stat)
res_CN_adapt_vs_stat <- results(dds_CN_adapt_vs_stat, contrast = c("group", "adaptation", "stationary"))
resultsNames(dds_CN_adapt_vs_stat)

# Log Fold Change Shrinkage with apeglm
resLFC_apeglm_CN_adapt_vs_stat <- lfcShrink(dds_CN_adapt_vs_stat, 
                                           coef = "group_adaptation_vs_stationary", 
                                           type = "apeglm")

# # Independent Hypothesis Weighting
# resIHW <- results(dds_CN_adapt_vs_stat, filterFun = ihw)
# summary(resIHW)
# sum(resIHW$padj < 0.05, na.rm = TRUE)
# metadata(resIHW)$ihwResult


# dir.create(paste0(pc_folders[pc_folder],
#                         "Dropbox/FHWN/M.BDSC.B.23.AA Masterarbeit/MTX/",
#                         enrichment,
#                         '/featurecounts/individual/',
#                         genome),
#                         recursive = TRUE, showWarnings = FALSE)

write.csv(as.data.frame(resLFC_apeglm_CN_adapt_vs_stat), 
          file = paste0(pc_folders[pc_folder],
                        "Dropbox/FHWN/M.BDSC.B.23.AA Masterarbeit/MTX/",
                        enrichment,
                        '/featurecounts/individual/',
                        genome,
                        "/dds_CN_adapt_vs_stat_resLFC_apeglm.csv"))


# MA Plots
plotMA(res_CN_adapt_vs_stat, ylim = c(-4, 9))
plotMA(resLFC_apeglm_CN_adapt_vs_stat, ylim = c(-4, 9))

# Dispersion Plot
plotDispEsts(dds_CN_adapt_vs_stat)

# Count Plots
plotCounts(dds_CN_adapt_vs_stat, gene = which.min(res$padj), intgroup = "group")
```


## Subset Data for "exponential" group
# Perform DESeq Analysis on Subset
# Log Fold Change Shrinkage with apeglm
# Independent Hypothesis Weighting
# Export Results
# MA Plots
# Dispersion Plot
# Count Plots

```{r subset-cn}
# Subset Data for CN Condition

dds_exponential <- dds[, dds$group == "exponential"]

dds_exponential$group <- droplevels(dds_exponential$group)
dds_exponential$condition <- droplevels(dds_exponential$condition)

dds_exponential$condition <- relevel(dds_exponential$condition, ref = "NH4")

design(dds_exponential) <- ~ condition

dds_exponential <- DESeq(dds_exponential)

resultsNames(dds_exponential)

res_exponential <- results(dds_exponential, contrast = c("condition", "CN", "NH4"))


# Log Fold Change Shrinkage with apeglm
resLFC_apeglm_exponential <- lfcShrink(dds_exponential, 
                                           coef = "condition_CN_vs_NH4", 
                                           type = "apeglm")

# # Independent Hypothesis Weighting
# resIHW <- results(dds_exponential, filterFun = ihw)
# summary(resIHW)
# sum(resIHW$padj < 0.05, na.rm = TRUE)
# metadata(resIHW)$ihwResult


# dir.create(paste0(pc_folders[pc_folder],
#                         "Dropbox/FHWN/M.BDSC.B.23.AA Masterarbeit/MTX/",
#                         enrichment,
#                         '/featurecounts/individual/',
#                         genome),
#                         recursive = TRUE, showWarnings = FALSE)

write.csv(as.data.frame(resLFC_apeglm_exponential), 
          file = paste0(pc_folders[pc_folder],
                        "Dropbox/FHWN/M.BDSC.B.23.AA Masterarbeit/MTX/",
                        enrichment,
                        '/featurecounts/individual/',
                        genome,
                        "/dds_exp_CN_vs_NH4_resLFC_apeglm.csv"))


# MA Plots
plotMA(res_exponential, ylim = c(-4, 9))
plotMA(resLFC_apeglm_exponential, ylim = c(-4, 9))

# Dispersion Plot
plotDispEsts(dds_exponential)

# Count Plots
plotCounts(dds_exponential, gene = which.min(res$padj), intgroup = "condition")

```

## Subset Data for "exponential" group
# Perform DESeq Analysis on Subset
# Log Fold Change Shrinkage with apeglm
# Independent Hypothesis Weighting
# Export Results
# MA Plots
# Dispersion Plot
# Count Plots

```{r subset-cn}
# Subset Data for CN Condition

dds_stationary <- dds[, dds$group == "stationary"]

dds_stationary$group <- droplevels(dds_stationary$group)
dds_stationary$condition <- droplevels(dds_stationary$condition)

dds_stationary$condition <- relevel(dds_stationary$condition, ref = "NH4")

design(dds_stationary) <- ~ condition

dds_stationary <- DESeq(dds_stationary)

resultsNames(dds_stationary)

res_stationary <- results(dds_stationary, contrast = c("condition", "CN", "NH4"))


# Log Fold Change Shrinkage with apeglm
resLFC_apeglm_stationary <- lfcShrink(dds_stationary, 
                                           coef = "condition_CN_vs_NH4", 
                                           type = "apeglm")

# # Independent Hypothesis Weighting
# resIHW <- results(dds_stationary, filterFun = ihw)
# summary(resIHW)
# sum(resIHW$padj < 0.05, na.rm = TRUE)
# metadata(resIHW)$ihwResult


# dir.create(paste0(pc_folders[pc_folder],
#                         "Dropbox/FHWN/M.BDSC.B.23.AA Masterarbeit/MTX/",
#                         enrichment,
#                         '/featurecounts/individual/',
#                         genome),
#                         recursive = TRUE, showWarnings = FALSE)

write.csv(as.data.frame(resLFC_apeglm_stationary), 
          file = paste0(pc_folders[pc_folder],
                        "Dropbox/FHWN/M.BDSC.B.23.AA Masterarbeit/MTX/",
                        enrichment,
                        '/featurecounts/individual/',
                        genome,
                        "/dds_stat_CN_vs_NH4_resLFC_apeglm.csv"))


# MA Plots
plotMA(res_stationary, ylim = c(-4, 9))
plotMA(resLFC_apeglm_stationary, ylim = c(-4, 9))

# Dispersion Plot
plotDispEsts(dds_stationary)

# Count Plots
plotCounts(dds_stationary, gene = which.min(res$padj), intgroup = "condition")

```